---
title: "Data transformation: dplyr package"
output: html_notebook
---


```{r xaringan-themer, include=FALSE, warning=FALSE}
# https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html#themes
library(xaringanthemer)
library(flextable)
library(kableExtra)
library(officer)
library(kableExtra)
library(rmarkdown)
library(openxlsx)
library(readxl)

std_border = fp_border(color="gray")
std_borderFoot = fp_border(color="black", width = 2)

```

# Introduction

Data transformation is an important step before starting to visualize and analyze your data. Transformation may involve:

* creating new variables and summaries
* renaming some variables
* reordering observations
* format variables
* select variables
* perform analyses by group
* filter/ subset observations
* etc ...

The *dplyr* package is the workhorse for data transformation in R.


# Installing dplyr

* as a standalone package:

_install.packages("dplyr")_

  * which is then called in R by:
  
      _library(dplyr)_

* as part of the *tidyverse* package

_install.packages('tidyverse')_

  * Note: *tidyverse* is a collection of packages for data science:
  
      * dplyr
      * ggplot2
      * tidyr 
      * ...
    
  * here, loading *tidyverse* automatically loads *dplyr* plus the rest of the packages
  
      _library(tidyverse)_

We will load *dplyr* as part of *tidyverse* package and show some of its uses on the iris data

```{r loaddplyr, echo=TRUE, warning=FALSE, message=FALSE, eval=TRUE}

#install.packages('tidyverse') # uncomment to install tidyverse

library(tidyverse) 

iris = iris

```

A subset of the iris dataset is shown in the table below:

```{r irishead, echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE, results='asis'}

flextable(head(iris)) %>% hline(part="body", border = std_border) %>% 
  hline_bottom(part="body", border = std_borderFoot) %>% 
  hline_top(part="body", border = std_borderFoot) %>% vline(border = std_border, part = "all") %>% 
  align(align = "center", part = "all") %>% align(align = "left", part = "all", j = 1) %>% 
  fix_border_issues(part = "all") %>% hline_top(part="header", border = std_borderFoot) %>% 
  fontsize(size = 12, part = "all")

```


In the next slides, we will show some key dplyr functions and their examples on the iris data

