---
title: "Sports data EPL2015-16"
format: html
  
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false


library(tidyverse)
library(flextable)
library(officer)

std_border = fp_border(color="gray")
std_borderFoot = fp_border(color="black", width = 2)

# minmax standardization

minmax_fn = function(x) (x - min(x))/(max(x) - min(x))

# format numeric cells

format_fn = function(x) formatC(x, format = 'f', digits = 2)

```



```{r,echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}

# <!-- https://statsbomb.com/wp-content/uploads/2022/08/Working-with-R.pdf -->

#| echo: false
#| eval: false
#| warning: false
#| message: false
#| output: false

# library(StatsBombR)
# 
# Comp = FreeCompetitions() |> 
#   filter(country_name == 'England' & competition_name == 'Premier League' &
#            competition_gender == 'male' & season_name == '2015/2016')
# 
# Matches = FreeMatches(Comp) 
# 
# StatsBombData <- free_allevents(MatchesDF = Matches, Parallel = T)
# StatsBombData = allclean(StatsBombData) 
# 
# # meaning of each column
# # 
# 
# # names(StatsBombData)
# 
# Matches =  Matches |> select("match_id", match_date, contains(c('score')), "referee.name", "referee.id", 
#          "home_team.home_team_id", "home_team.home_team_name", "away_team.away_team_id", 
#          "away_team.away_team_name") 
# 
# write.csv(Matches, "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/Matches.csv", 
# row.names = FALSE)
# 
# # shots per match per team
# # 
# 
# dfshots = StatsBombData %>% filter(type.name == 'Shot') |> 
#   select(match_id, possession_team.name, possession_team.id, type.name) %>% 
#   group_by(match_id, possession_team.id, possession_team.name) %>% count()
#   
# write.csv(dfshots, "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfshots.csv", 
# row.names = FALSE)
# 
# # tactics applied by each team for a match ID
# #   
# 

# dftactics = StatsBombData |> filter(!is.na(tactics.formation) & type.name=='Starting XI') %>%
#   select(match_id, team.name, team.id, tactics.formation)

# 
# write.csv(dftactics, "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dftactics.csv", 
# row.names = FALSE)
# 
# # player-position
# 
# write.csv(unique(StatsBombData |> select(position.id, position.name)), 
#           "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfplayerpositions.csv",
#           row.names = FALSE)

# cards received per match

# View(StatsBombData |> filter(!is.na(foul_committed.card.name)) |> select(match_id, team.name, possession_team.name, possession, bad_behaviour.card.name, foul_committed.card.name))

# write.csv(StatsBombData |> filter(!is.na(foul_committed.card.name)) |>
# group_by(match_id, team.name, team.id) |> count(),
# "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfcards.csv",
# row.names = FALSE)


```

# Variables

Dependent variables

- home/away team goals (Matches data)
- shots on target (depending on team, home/away etc)

Important independent variables:

-   away or home team (matches data)
-   position id/name: i.e the position of the player

```{r,echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}


#| tbl-cap-location: top
#| tbl-cap: "positions of players"
#| 
#| 

dfplayerpositions = read.csv(
  "C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfplayerpositions.csv")

flextable(dfplayerpositions) |> 
  hline(part="body", border = std_border) |> 
  hline_bottom(part="body", border = std_borderFoot) |> 
  hline_top(part="body", border = std_borderFoot) |> 
  merge_h(part = "header") |> merge_v(part = "header") |> 
  vline(border = std_border, part = "all") |> align(align = "center", part = "all") |> 
  align(align = "left", part = "all", j = 1) |> width(j = c(1:2), width = c(0.75, 1.5)) |>
  fix_border_issues(part = "all") |> hline_top(part="header", border = std_borderFoot) |> 
  fontsize(size = 10, part = "all")


```

-   fouls committed
-   visits to opposition half
-   tactics formation at start of match
-   number of corners

# Questions

* probability of winning as a function of 

  * shots per match, cards received per match, home or away team, tactics employed
  
# Head of win-lose data

```{r }
#| echo: false
#| warning: false
#| message: false


dfcards = read.csv("C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfcards.csv") |>
  rename(ncards = `n`) |> select(-`team.name`)

dfshots = read.csv("C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dfshots.csv") |>
  rename(team.id = `possession_team.id`, nshots = `n`) |> select(-`possession_team.name`)

dftactics = read.csv("C:/Users/Public/lmaaya/data/Sports data EPL2016-16/dftactics.csv") |>
  rename(formation = `tactics.formation`) |> select(-`team.name`)

Matches = read.csv("C:/Users/Public/lmaaya/data/Sports data EPL2016-16/Matches.csv")

# Independent factors: away or home team, shots, cards, tactics, referee
# dependent factor : win or loose, draw?
# mutate(Winordraw = ifelse(home_score == away_score, 0, 1)) |>

dfMatches = Matches |> select(-contains(c('name', 'date', 'referee'))) |>
  filter(home_score != away_score) |> # only where there's a winner
  mutate(home_win = ifelse(home_score> away_score, 1, 0), 
         away_win = ifelse(home_score< away_score, 1, 0)) |>
  rename(home_team = `home_team.home_team_id`, away_team = `away_team.away_team_id`) |>
  pivot_longer(!1, names_to = c("hometeam", ".value"), names_pattern = "(.+)_(.+)") |>
  rename(team.id = `team`) |> mutate(hometeam = as.factor(ifelse(hometeam=='home', 1, 0)))

dfMatches = left_join(dfMatches, dftactics, by = c("match_id", "team.id")) |>
  left_join(dfshots, by = c("match_id", "team.id")) |> 
  left_join(dfcards, by = c("match_id", "team.id")) |> 
  mutate(ncards = ifelse(is.na(ncards), 0, ncards))


dfMatchesTab = as.data.frame(dfMatches |> mutate_if(is_integer, as.character)) |>
  slice(1:6)

flextable(dfMatchesTab) |> 
  hline(part="body", border = std_border) |> 
  hline_bottom(part="body", border = std_borderFoot) |> 
  hline_top(part="body", border = std_borderFoot) |> 
  merge_h(part = "header") |> merge_v(part = "header") |> 
  vline(border = std_border, part = "all") |> align(align = "center", part = "all") |> 
  fix_border_issues(part = "all") |> hline_top(part="header", border = std_borderFoot) |> 
  fontsize(size = 10, part = "all")

```

# fit model for probability of winning

```{r}
#| echo: false
#| warning: false
#| message: false

cols = c('nshots', 'ncards')

dfMatches = dfMatches |> mutate(formation = as.factor(formation)) |> 
  mutate_at(.vars = cols, .funs = minmax_fn)

model <- glm(win ~ hometeam + formation + nshots + ncards, family="binomial", 
             data = dfMatches)

results = summary(model)['coefficients'] |> as.data.frame() |> rownames_to_column() |>
  mutate_if(is.numeric, .funs = format_fn) |> 
  rename('term' = `rowname`, 'estimate' = 2, 'sderror' = 3, 'z-value' = 4, 'p-value' = 5)


flextable(results) |> 
  hline(part="body", border = std_border) |> 
  hline_bottom(part="body", border = std_borderFoot) |> 
  hline_top(part="body", border = std_borderFoot) |> 
  merge_h(part = "header") |> merge_v(part = "header") |> 
  vline(border = std_border, part = "all") |> align(align = "center", part = "all") |> 
  fix_border_issues(part = "all") |> hline_top(part="header", border = std_borderFoot) |> 
  fontsize(size = 10, part = "all")

```

